Place of death in NRS COVID-19 deaths

### 2021-01-25 covid_nrs_place_of_death.Rmd
added past deaths so we can compute averages, min & max by week number. started explaining how some years have 52 and some 53 weeks

```{r}
# checking out structure
ods_structure("deaths-involving-coronavirus-covid-19")
ods_schemes("deaths-involving-coronavirus-covid-19")
ods_print_query("deaths-involving-coronavirus-covid-19")

@@

# TODO: remove below bit, it's for comparison with previous results!
cv_data_raw <- read_csv("https://statistics.gov.scot/downloads/cube-table?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19")
cv_data_main <- cv_data_raw %>% 
  clean_names() %>% 
  filter(feature_code == "S92000003") %>% 
  mutate(date_code = str_replace(date_code, "w/c ", ""),
         date_code = ymd(date_code)) %>% 
  filter(!is.na(date_code)) %>% 
  select(date_code, sex, age, cause_of_death, location_of_death, measurement, value, 
         -feature_code, -units) %>% 
  arrange(date_code, sex, age, cause_of_death, location_of_death, measurement)
```

@@

@@

##

>>>>

```{r}
# ods_all_datasets() %>% View()
# ods_structure("deaths")

# checking out structure
ods_structure("deaths-involving-coronavirus-covid-19")
ods_schemes("deaths-involving-coronavirus-covid-19")
ods_print_query("deaths-involving-coronavirus-covid-19")

@@

# TODO: remove below bit, it's for comparison with previous results!
cv_data_raw <- read_csv("https://statistics.gov.scot/downloads/cube-table?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19")
cv_data_main <- cv_data_raw %>% 
  clean_names() %>% 
  filter(feature_code == "S92000003") %>% 
  mutate(date_code = str_replace(date_code, "w/c ", ""),
         date_code = ymd(date_code)) %>% 
  filter(!is.na(date_code)) %>% 
  select(date_code, sex, age, cause_of_death, location_of_death, measurement, value, 
         -feature_code, -units) %>% 
  arrange(date_code, sex, age, cause_of_death, location_of_death, measurement)

## load past deaths for calculating averages
temp_file <- tempfile()
past_deaths_source <- "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv"
past_deaths_file <- curl_download(url = past_deaths_source, destfile = temp_file, quiet = FALSE)

past_deaths <- 
  read_csv(past_deaths_file, skip = 2) %>%
  rename(year_of_death = 1) %>%  # rename first column
  mutate(
    place_of_death = if_else(is.na(`1`), true = year_of_death, NA_character_)
  ) %>%
  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
  drop_na(`1`) %>%
  pivot_longer(cols = all_of(as.character(1:53)), values_to = "value", names_to = "week_number") %>%
  mutate(week_number=as.integer(week_number))

past_deaths %>% filter(week_number %in% 52:53) %>% count(year_of_death, week_number)

past_deaths_summarised <-
  past_deaths %>%
  filter(year_of_death!="average") %>%  # remove averages
  group_by(year_of_death, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
  filter(!all(value==0 & week_number==53)) %>%
  ungroup %>%
  group_by(week_number,place_of_death) %>%
  summarise(across(value, .fns = list(mean=mean,min=min,max=max)),  # having removed 0s for week 53 where it didn't exist, we average over years for week 53 that had that week only
    .groups = "drop"
  ) %>% rename_all(~str_replace_all(.x, pattern = "value", replacement = "deaths"))

# TODO: show that only 2015 had week 53 and deal with this somehow - combine next years week 1 and previous week 52 instead?
week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks."
```

@@

## Recoding place of death

```{r}
## recoding place of death
data_main %>% count(location_of_death)
past_deaths %>% count(place_of_death)

# datapasta::tribble_paste(input_table = tibble(location_of_death = sort(unique(data_main$location_of_death)), place_of_death = sort(unique(past_deaths$place_of_death))))
tbl_recoded_place_of_death <- tibble::tribble(
  ~location_of_death,           ~place_of_death,   ~place_of_death_recoded,
  "all",                        "All locations",   "All",
  "care-home",                      "Care Home",   "Care home",
  "-non-institution", "Home and other non-institution",   "Home",
  "hospital",                       "Hospital",   "Hospital",
  "other-institution",    "Other institution3",   "Other"
     )
```

## Merged data

```{r}
## check whether there's an issue with week 52/53
#

merged_data <-
  data_main %>%
  left_join(tbl_recoded_place_of_death) %>%
  left_join(past_deaths %>% mutate(sex = "all", age = "all", cause_of_death = "all-causes"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death", "week_number", "sex", "age", "cause_of_death")) %>% 
  select(-c(location_of_death,place_of_death))

```

@@

## Deaths over time

```{r, fig.width=12, fig.height=9}

data_main %>% 
  # filter(sex == "all" & location_of_death == "all" & cause_of_death == "covid-19-related") %>% 
  # filter(age != "all") %>% 
  group_by(age) %>% 
  summarise(n = sum(value)) %>%
  mutate(pct = round(n/sum(n), 3)) %>% 
  ggplot(aes(age, n)) +
  geom_col(fill = "#4477AA") +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80)) +
  theme_minimal() +
  labs(x = "",
       y = "",
       title = "Deaths Associated with COVID-19 by Age Group",
       subtitle = date_range_text,
       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate.")
```
